Pool Model Paper

### 2023-08-23 pool_model.tex
some mistakes correction

\section{General Mathemetical Formulation}
Consider the system conssisting of $N$ species in $M$ different states.
In the general representation the system can be described with the Ordinary Differential Equations (ODEs) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In a matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_M  \\
        \vdots &       & \vdots \\
        x^N_1  & \dots & x^N_M  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{ij}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu, ...).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

The advantage of the following formulation is its generality. 
Depending on the knowledge about system as well as the needed level of accuracy one can define arbitrary number of states $M$ and control the complexity of teh model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.

@@

Due to the life in rather unfavourable conditions bacteria developed different surviving mechanisms, one of which is adapting their metabolism and transitioning into dormant state.
By defeinition dormancy in bacteria is a reversible state where cell continue living with low metabolic activity while the cell division stops~\cite{kaprelyants_dormancy_1993}. 
This survival strategy can appear in response to different external stresses, e.g., nutritient deficiency, temperature stress or extreme fluctuations of environment parameters.
And after "difficult times" bacteria, also called viable but non-culturable(VBNC), can continue cell division~\cite{kell_viability_1998}.
The models including transition of bacteria to the dormant state were proposed and studied by Bär \etal~\cite{bar_modelling_2002} and Jones and Lennon~\cite{jones_dormancy_2010}.

@@

One of the reasons of the dormant state appearence can be also the temperature shift~\cite{oliver_viable_1995}.

We propose a kind of visco-elastic Maxwell type of stress-strain relation and write down the ordinary differential equation for $\gamma$ to be:
\begin{eqnarray}
\dot{\gamma} &=& \Gamma \left |\dot{T}\right |-\delta \gamma.
\end{eqnarray}
We assumed that the direction of the temperature change does not matter, i.e., a change from $T=2$ to $T=12$ has the same effect as a change from $T=14$ to $T=4$. 
$\Gamma$ is a scaling factor and $\delta$ is the relaxation time for the temperature disturbance. 
Integration of this equation yields (with $\gamma(0)=0$):
\begin{eqnarray}
\gamma(t) &=& \Gamma \int_0^t \left |\dot{T}(t^{\prime})\right |e^{-\delta (t-t^{\prime})}dt^{\prime}.
\end{eqnarray}
After some manipulation of this equation we find:
\begin{eqnarray}
\gamma(t) &=& \Gamma\delta\left |\left[T[t]-T[0]e^{-\delta t}-\delta \int_0^t T(t^{\prime})e^{-\delta (t-t^{\prime})}dt^{\prime}\right]\right |.
\end{eqnarray}
This representation of the rate $\gamma(t)$ has the advantage that it only involves $T(t)$ and does not require the calculation of the derivative of $T$. 
If $T$ exhibits $n$ temperature jumps, one finds:
\begin{eqnarray}
\gamma(t) &=& \Gamma\sum_{i=1}^n \left |\Delta T_i \right |\theta(t-t_i)e^{-\delta(t-t_i)}.
\end{eqnarray}

>>>>

\section{General Mathemetical Formulation}
Consider the system conssisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (ODEs) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_M  \\
        \vdots &       & \vdots \\
        x^N_1  & \dots & x^N_M  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{ij}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu, ...).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

The advantage of the following formulation is its generality. 
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.

@@

Due to its life in rather unfavourable conditions bacteria developed different surviving mechanisms, one of which is adapting their metabolism and transitioning into dormant state.
By defeinition dormancy in bacteria is a reversible state where cell continue living with low metabolic activity while the cell division stops~\cite{kaprelyants_dormancy_1993}. 
This survival strategy can appear in response to different external stresses, e.g., nutritient deficiency, temperature stress or extreme fluctuations of environment parameters.
And after "difficult times" bacteria, also called viable but non-culturable (VBNC), can continue cell division~\cite{kell_viability_1998}.
The models including transition of bacteria to the dormant state were proposed and studied by Bär \etal~\cite{bar_modelling_2002} and Jones and Lennon~\cite{jones_dormancy_2010}.

@@

One of the reasons of the dormant state appearence can be also the temperature shift~\cite{oliver_viable_1995}.
Moreover, the rapid shift in temperature is associated with the appearence of an additional lag phase~\cite{zwietering_modeling_1994}.
To model it we propose a kind of visco-elastic Maxwell type of stress-strain relation and write down the ordinary differential equation for $\gamma$ to be:
\begin{eqnarray}
\dot{\gamma} &=& \Gamma \left |\dot{T}\right |-\delta \gamma.
\end{eqnarray}
We assumed that the direction of the temperature change does not matter, i.e., a change from $T=2$ to $T=12$ has the same effect as a change from $T=14$ to $T=4$. 
$\Gamma$ is a scaling factor and $\delta$ is the relaxation time for the temperature disturbance. 
Integration of this equation yields (with $\gamma(0)=0$):
\begin{eqnarray}
\gamma(t) &=& \Gamma \int_0^t \left |\dot{T}(t^{\prime})\right |e^{-\delta (t-t^{\prime})}dt^{\prime}.
\end{eqnarray}
After some manipulation of this equation we find:
\begin{eqnarray}
\gamma(t) &=& \Gamma\delta\left |\left[T[t]-T[0]e^{-\delta t}-\delta \int_0^t T(t^{\prime})e^{-\delta (t-t^{\prime})}dt^{\prime}\right]\right |.
\end{eqnarray}
This representation of the rate $\gamma(t)$ has the advantage that it only involves $T(t)$ and does not require the calculation of the derivative of $T$. 
If $T$ exhibits $n$ temperature jumps, one finds:
\begin{eqnarray}
\gamma(t) &=& \Gamma\sum_{i=1}^n \left |\Delta T_i \right |\theta(t-t_i)e^{-\delta(t-t_i)}.
\end{eqnarray}
